---
title: "Case Study 1 - Cyclistic"
author: "Kyle Hollands"
date: "10/03/2022"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Table of Contents

- Objective
- What the data shows
- Conclusion

## Objectives

- How do annual members and casual riders use Cyclistic bikes differently?

## What the data shows

- Casual riders use the service more on weekends; less during the week.
- Members use the service consistently throughout the entire week.
- Casual rides appear to be longer on average than member rides.

```{r scripts, echo = FALSE, include=FALSE,message=FALSE, warning=FALSE}
# Load the required libraries
library(tidyverse)
library(janitor)
library(readr)
library(tidyr)
library(chron)
library(dplyr)
library(readr)
library(lubridate)

options("scipen"=10,
dplyr.summarise.inform = FALSE)

# Import and combine twelve months of data into one dataset
combined_trip_data <- list.files(path = "./Datasets/Modified CSV Files",  # Identify all CSV files
  pattern = "*.csv", full.names = TRUE) %>% 
  lapply(read_csv) %>%                              # Store all files in list
  bind_rows                                         # Combine data sets into one data set

# Rename columns to be more readable and self-explanatory
(combined_trip_data <- rename(combined_trip_data
                   ,trip_id = ride_id
                   ,start_time = started_at
                   ,end_time = ended_at
                   ,from_station_name = start_station_name
                   ,from_station_id = start_station_id
                   ,to_station_name = end_station_name
                   ,to_station_id = end_station_id
                   ,usertype = member_casual))

# Trim the dataset to only include the most relevant information
trimmed_trip_data <- combined_trip_data %>%
  select(-c(start_lat, start_lng, end_lat, end_lng, from_station_name,from_station_id,to_station_name,to_station_id))

# Inspect the data
colnames(trimmed_trip_data)
glimpse(trimmed_trip_data)
nrow(trimmed_trip_data)
dim(trimmed_trip_data)
head(trimmed_trip_data)
tail(trimmed_trip_data)
summary(trimmed_trip_data)
str(trimmed_trip_data)

trimmed_trip_data$date <- as.Date(trimmed_trip_data$start_time) #The default format is yyyy-mm-dd
trimmed_trip_data$month <- format(as.Date(trimmed_trip_data$date), "%m")
trimmed_trip_data$day <- format(as.Date(trimmed_trip_data$date), "%d")
trimmed_trip_data$year <- format(as.Date(trimmed_trip_data$date), "%Y")

# Recode the day of week from a number to a day. (Data was initially manipulated calculating
# day of week as a double.)
trimmed_trip_data$day_of_week <- format(as.Date(trimmed_trip_data$date), "%A")

# Determine which columns/rows contain the most NA's.
sapply(trimmed_trip_data, function(x) sum(is.na(x)))

# Remove any rows where ride length <= 0 seconds and greater than 3 hours.
cleaned_trip_data <- trimmed_trip_data[!(trimmed_trip_data$ride_length <= 0 | trimmed_trip_data$ride_length >= 10800),]
sapply(cleaned_trip_data, function(x) sum(is.na(x)))

# Remove rows with NAs.
cleaned_trip_data <- na.omit(cleaned_trip_data)
```

## Ride Statistics
```{r statistics_2, echo=FALSE, fig.keep='all', warning=FALSE, message=FALSE, tidy=TRUE}
cleaned_trip_data %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>% 
  summarise(number_of_rides = n(), 
  average_duration = mean(ride_length)) %>%
  arrange(usertype, weekday)
```

##
```{r chart_2, echo=FALSE,results='hide',fig.keep='all',warning=FALSE,message=FALSE}
cleaned_trip_data %>%
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>%
  summarise(number_of_rides = n(),
  average_duration = mean(ride_length)) %>%
  arrange(usertype, weekday) %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = usertype)) +
  geom_col(position = "dodge") +
  labs(
    x = "Day of Week", y = "Average Ride Length",
    color = "New Legend Title",
    title = "Average ride length per usertype by day of week",
    subtitle = "Estimates based on data collected over the past twelve months.",
    caption = "Source: Cyclistic bikeshare terminals"
  ) +
  scale_fill_discrete(name="Usertype") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

##
```{r chart_1, echo=FALSE, results='hide',fig.keep='all',warning=FALSE,message=FALSE}
cleaned_trip_data %>% 
  mutate(weekday = wday(start_time, label = TRUE)) %>% 
  group_by(usertype, weekday) %>%
  summarise(number_of_rides = n(),
  average_duration = mean(ride_length)) %>% 
  arrange(usertype, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge") +
  labs(
    x = "Day of Week", y = "# of Rides",
    color = NULL,
    title = "Number of rides each day by usertype",
    subtitle = "Estimates based on data collected over the past twelve months.",
    caption = "Source: Cyclistic bikeshare terminals"
  ) +
  scale_fill_discrete(name="Usertype") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Conclusion

- Casual riders are using the service, on average, longer than members and appear
to be using them for more leisure activities. (Higher usage on weekends.)

- Members are using the service more consistently throughout the week for
shorter durations, which could indicate they are using it as a form of transit
to and from work.

## Next Steps

- Determine how to convert casual riders into members.
  - More data will be required for further analysis.
